PERs from Partial Projections for Higher-order Polymorphic Demand Analysis

### 2017-02-22 pubs/demand_analysis/new.verb.tex
test push

\begin{abstract}
  {\it First-order\/} projection-based strictness analysis (properly,
  \emph{demand analysis}) was first introduced by Wadler and Hughes in
  1987~\cite{WH87}.  This development was restricted to first-order,
  monomorphic languages and provided no proof of soundness.  Subsequent
  developments extended the theory to other analyses, approaches to higher
  order, and polymorphism.  Yet other developments were based on
  generalization of projections, specifically partial equivalence relations
  and partial projections, the latter notably simplifying a difficulty with
  projection-based strictness dating back to Wadler and Hughes.  However, to
  date no development has combined the desirable contributions of all of these
  approaches.  We present the first practical higher-order, polymorphic,
  partial-projection/partial-equivalence-relation based demand analysis,
  together with a proof of satisfaction of the fundamental safety condition
  for such analyses.  Such an analysis can be used both to improve serial code
  generation, as well as enable safe automatic parallelization, for lazy or
  non-strict functional languages.

@@

We take as given that demand analysis is essential for good serial code
generation, and for identifying where parallelism may be safely introduced.
Numerous such analysis techniques have been proposed and implemented; we
greatly narrow the field of discussion by restricting attention to those for
which there is a formally stated notion of correctness that the technique has
been proven to satisfy.

Analysis techniques can usually be identified as being based on either
a non-standard denotational semantics or a non-standard typing.
Examples in the latter category include those of Gomard \cite{Go92},
Jensen \cite{Jen92}, the Nielsons \cite{NN88}, Schmidt \cite{Sch88},
and Henglein and Mossin \cite{HM94}.  Our focus is on those techniques
based on non-standard interpretation, in particular, those using
projections or partial equivalence relations (PERs) as the basic
abstract values.

A domain {\it projection\/} is a continuous idempotent function that
approximates the identity.  Launchbury \cite{Lau88} hit upon the idea
of using projections to encode degrees of staticness of data.  The
basic idea is that a projection maps to $\bot$ that part of a data
structure that is dynamic (possibly not determined), and acts as the
identity on that part which is static (definitely determined).
Examples are the identity $\ID$, the greatest projection, which
specifies that values are entirely static; the constant $\bot$
function $\BOT$, the least projection, which specifies that values are
entirely dynamic; and projections $\FST$ and $\SND$ on product
domains, defined by
\[
\FST\ (x,y) = (x,\bot)\ ,\hspace{1.0in} \SND\ (x,y) = (\bot,y)\ ,
\]
specifying staticness in the first and second components of pairs,
respectively.  The nominal goal of analysis is, given function $f$
denoted by some programming-language expression, and projection $\D$
encoding the staticness of the argument of $f$, to determine $\G$
satisfying the {\it safety condition\/} $\G\o\f\we\f\o\D$.  For
example, taking $\G$ to be $\SND$ satisfies
$\G\o\itt{swap}\we\itt{swap}\o\FST$ for $\itt{swap}$ defined by
$\itt{swap}~(x,y)=(y,x)$.  Taking $\G$ to be $\BOT$ always satisfies
the safety condition but tells nothing; greater $\G$ is more
informative.  Launchbury \cite{Lau91a} showed that this safety
condition, in a sense which he formalises, is equivalent
to the correctness condition for binding-time analysis in the general
framework of Jones \cite{Jon88}.  Using projection-based analysis,
Launchbury implemented both monomorphic and polymorphic versions of a
partial evaluator for a first-order language.

>>>>

\begin{abstract}
  {\it First-order\/} projection-based strictness analysis (generally,
  \emph{demand analysis}) was first introduced by Wadler and Hughes in
  1987~\cite{WH87}.  This development was restricted to first-order,
  monomorphic languages and provided no proof of soundness.  Subsequent
  developments extended the theory to other analyses (most notably
  binding-time analysis), approaches to higher order, and polymorphism.  Yet
  other developments were based on generalization of projections, specifically
  partial equivalence relations and partial projections, the latter notably
  simplifying a difficulty with projection-based strictness analysis dating
  back to Wadler and Hughes.  However, to date no development has combined the
  desirable attributes of all of these approaches.  We present the first
  practical, higher-order, polymorphic,
  partial-projection/partial-equivalence-relation based demand analysis,
  together with a proof of satisfaction of the fundamental safety condition
  for such analyses.  Such an analysis can be used both to improve serial code
  generation as well as enable safe automatic parallelization for lazy or
  non-strict functional languages.

@@

We take as given that demand analysis, aka strictness analysis, is essential
for good serial code generation, and for identifying where parallelism may be
safely introduced.  An example of the former is the current ghc Haskell
compiler, where strictness analysis is used to unbox values and thereby
registerize code.  A potential application of the latter is in the context of
the APPFL parallelizing, which assumes strict semantics by default for the
purpose of exploring automatic parallelization techniques, but in the presence
of non-strict semantics could use such an analysis to enable such
parallelization.  Numerous such analysis techniques have been proposed and
implemented; we greatly narrow the field of discussion by restricting
attention to those for which there is a formally stated notion of correctness
that the technique has been proven to satisfy.

Analysis techniques can usually be identified as being based on either a
non-standard denotational semantics or a non-standard typing.  Examples in the
latter category include those of Gomard \cite{Go92}, Jensen \cite{Jen92}, the
Nielsons \cite{NN88}, Schmidt \cite{Sch88}, and Henglein and Mossin
\cite{HM94}.  Our focus is on those techniques based on non-standard
interpretation, in particular, those using projections, partial projections, or
partial equivalence relations (PERs) as the basic abstract values.

A domain {\it projection\/} is a continuous idempotent function that
approximates the identity.  Wadler and Hughes first presented the idea of
projection-based program analysis in the context of

Launchbury \cite{Lau88} hit upon the idea
of using projections to encode degrees of staticness of data.  The
basic idea is that a projection maps to $\bot$ that part of a data
structure that is dynamic (possibly not determined), and acts as the
identity on that part which is static (definitely determined).
Examples are the identity $\ID$, the greatest projection, which
specifies that values are entirely static; the constant $\bot$
function $\BOT$, the least projection, which specifies that values are
entirely dynamic; and projections $\FST$ and $\SND$ on product
domains, defined by
\[
\FST\ (x,y) = (x,\bot)\ ,\hspace{1.0in} \SND\ (x,y) = (\bot,y)\ ,
\]
specifying staticness in the first and second components of pairs,
respectively.  The nominal goal of analysis is, given function $f$
denoted by some programming-language expression, and projection $\D$
encoding the staticness of the argument of $f$, to determine $\G$
satisfying the {\it safety condition\/} $\G\o\f\we\f\o\D$.  For
example, taking $\G$ to be $\SND$ satisfies
$\G\o\itt{swap}\we\itt{swap}\o\FST$ for $\itt{swap}$ defined by
$\itt{swap}~(x,y)=(y,x)$.  Taking $\G$ to be $\BOT$ always satisfies
the safety condition but tells nothing; greater $\G$ is more
informative.  Launchbury \cite{Lau91a} showed that this safety
condition, in a sense which he formalises, is equivalent
to the correctness condition for binding-time analysis in the general
framework of Jones \cite{Jon88}.  Using projection-based analysis,
Launchbury implemented both monomorphic and polymorphic versions of a
partial evaluator for a first-order language.
